---
title: "hw4-irene-liang"
author: "Irene Liang"
date: "2022-10-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, error = TRUE)
library(tidyverse)
library(tidyr)
library(dplyr)
library(plotly)
```

# 1) Future Value (with compound interest)

## 1.1) Function future_value()

The Future Value (FV) with compound interest is given by:

$$
\begin{equation} \text{FV} = \text{P}\left(1+\frac{r}{k}\right)^{nk} \end{equation}
$$

```{r}
#' @title Future Value
#' @description computes the future value of an investment
#' @param principal initial amount
#' @param rate annual interest rate (or annual rate of return)
#' @param n number of years
#' @param k number of compounding periods per year
#' @return future value of investment
future_value <- function(principal = 100, r = 0.01, n = 1, k = 1){
  principal * (1 + r/k)^(n*k)
}
```

## 1.2) Test out the future_value() function

```{r}
future_value(principal = 1000, r = 0.025, n = 3, k = 1)
```

# 2) Future Value of Ordinary Annuity

The Future Value of an Ordinary Annuity is given by:

$$
\begin{equation} \text{FV} = \text{PMT} \times \frac{ \left(1+\frac{r}{k}\right)^{nk} - 1}{r/k}\end{equation}
$$ 

Notice that if r = 0 then:
$$
\begin{equation} \text{FV} = \text{PMT} \times nk \end{equation}
$$

## 2.1) Function annuity()

```{r}
#' @title Future Value of Ordinary Annuity
#' @description Computes the future value of an ordinary annuity
#' @param pmt periodic contribution made at the end of each period
#' @param r annual interest rate (i.e. annual rate of return)
#' @param n number of years
#' @param k number of compounding periods per year
#' @return future value of ordinary annuity

annuity <- function(pmt = 100, r = 0.01, n = 1, k = 12){
  if (r==0){
    fv_annuity <- pmt*n*k
  }
  else fv_annuity <- pmt * ((1 + r/k)^(n*k) - 1) / (r / k)
  return(fv_annuity)
}
```

## 2.2) Use your function annuity() to find how much money Mrs. Savingswood will have at the end of 10 years.

```{r}
annuity(pmt = 100, r = 0.05, n = 10, k = 12)
```

# 3) Combo: Future Value and Ordinary Annuity

## 3.1) Investment table for a 10-year period

```{r}
investment_over_yr <- vector(mode="numeric", length = 11)
for(n in 1:11){
  combo <- round(future_value(principal = 1000, r = 0.0755, n = n-1, k = 1) + 
                   annuity(pmt = 720, r = 0.0755, n=n-1, k = 1), digits = 2)
  investment_over_yr[n] <- combo
}

investment_10yr <- data.frame(year = 0:10,
                              amount = investment_over_yr)
investment_10yr
```

# 4) Simulating Investing in a Total Stock Market index fund

## 4.1) 50 simulations of a 10-year investment period

```{r}
initial_amount = 1000
num_years = 10
num_simulations = 50
annual_return = 0.1
annual_volatility = 0.18
set.seed(12345)

# initialize matrix to store outputs 
balances = matrix(0, nrow = num_years + 1, ncol = num_simulations)

for (sim in 1:num_simulations) {
  aux_balance = rep(initial_amount, num_years + 1) # initialize vector to store outputs 
  return_rates = rnorm(num_years, mean = annual_return, sd = annual_volatility)
  for (year in 1:num_years) {
    aux_balance[year + 1] = aux_balance[year] * (1 + return_rates[year]) + 360 # compound previous balance for one year
    }
  balances[ ,sim] = aux_balance
}

rownames(balances) = 0:num_years
colnames(balances) = paste0("sim", 1:num_simulations)
names(dimnames(balances)) <- c("year", "")
balances[, 1:3] # Display the first 3 columns of the balance matrix
```

## 4.2) Data Visualization

```{r}
# convert to data frame
balances_df = as.data.frame(balances)
balances_df <- balances_df %>% rowwise() %>% mutate(percentile25 = quantile(c_across(sim1:sim50), probs = 0.25),
                                                     percentile50 = quantile(c_across(sim1:sim50), probs = 0.50),
                                                     percentile75 = quantile(c_across(sim1:sim50), probs = 0.75))
balances_df$year = 0:(nrow(balances)-1)

# reshape table into "tall" or "long" format
balances_df_tall = pivot_longer(
  balances_df, 
  cols = starts_with(c("sim", "percentile")),
  names_to = "simulation_percentile", 
  values_to = "amount_stat")

# indicator for whether the row is a simulation
balances_df_tall$indicator <- str_detect(balances_df_tall$simulation_percentile, 'sim') 

timelines = ggplot(data = balances_df_tall, aes(x = year, y = amount_stat, group = simulation_percentile, 
                                                color = indicator, alpha = indicator), show.legend = FALSE) + 
  geom_line() +
  theme_minimal() +
  scale_alpha_manual(values = c('TRUE' = 0.1, 'FALSE' = 1)) +
  guides(color = 'none', alpha = 'none') +
  scale_color_manual(values = c("red", "black")) + 
  annotate(geom="text", label="25th Percentile", x=9.6, y=7000, color = "red", size = 3) +
  annotate(geom="text", label="50th Percentile", x=9.6, y=8800, color = "red", size = 3) +
  annotate(geom="text", label="75th Percentile", x=9.6, y=11200, color = "red", size = 3) +
  theme(legend.position='none') +
  labs(y = "balance ($)", x = "years") +
  ggtitle("Timeline for 50 Simulations of a 10-year Investment Period")

ggplotly(timelines)
```
